---
title: "Homework 1"
author: "Sergio Ozoria Ramírez (smo2147)"
date: "2025-09-18"
output: github_document
---

# Initial Setup

Loading the tidyverse package for scatterplot + moderndive package to access the penguins dataset.

```{r initial setup, message = FALSE}
library(tidyverse)
library(moderndive)
```

# Problem 1

#### Description of Dataset

We will now access the penguins dataset and provide a short description of the data using inline R code.

```{r describing dataset}
data("early_january_weather")
```

The penguins dataset has **`r nrow(early_january_weather)` rows** and **`r ncol(early_january_weather)` columns**. One important variable that stands out in this dataset is the **temperature** variable (i.e., `temp`), along with hour (i.e., `hour`) and the wind speed (i.e., `wind_speed`), which are critical in regulating/influencing temperature levels. It is important to note that the average temperature has a mean of **`r early_january_weather |> pull(temp) |> mean()`** in Fahrenheit. Moreover, the average hour is **`r early_january_weather |> pull(hour) |> mean()`** and the average wind speed is **`r early_january_weather |> pull(wind_speed) |> mean()`** in the month of January, 2023.

The following table describes the remaining variables that were also used to capture observations in weather patterns in the month of January, year 2023:

Variable   | Description
---------- | -----------
`origin`   | Location where the weather observation was recorded
`year`     | Year of observation
`month`    | Month of observation
`hour`     | Hour of the day
`day`      | Day of the month of observation
`dewp`     | Dew point temperature, indicating moisture in air
`humid`    | Humidity at time of observation
`wind_dir` | Wind direction at time of observation
`wind_gust`| Peak instantaneous wind speed at observation
`precip`   | Precipiation at the time of observation
`pressure` | Atmospheric pressure at time of observation
`visib`    | Visibility at time of observation
`time_hour`| Time and hour at the time of observation

#### Dataframe for Scatterplot

Before we plot what the temperature by the hour looks like in a scattersplot, let's use the selection function to create a dataframe for the variables `time_hour`, `temp`, and `humid`.

```{r creating dataframe}
weather_df =
  early_january_weather |>
  select(time_hour, temp, humid
  )
```

#### Plotting Scatterplot

We will now plot the variables `temp` and `time_hour` to see how the temperature changed by the hour.

```{r plotting weather patterns}
ggplot(
  weather_df, aes
       (x = time_hour, 
        y = temp, 
        color = humid)) + 
  geom_point() + 
  labs(
    x = "Hour of the Day", 
    y = "Temperature (°F)", 
    color = "Humidity (%)") +
  ggtitle("Temperature Patterns by Time and Hour in January, 2023") +
  theme(plot.title 
        = element_text
        (hjust = 0.5)
      )
```

This scatterplot depicts temperature patterns over time, as well as changes in humidity levels throughout the month of January in 2023. As the temperature rises by the hour over the course of the month, humidity levels also tend to increase. It is important to note that humidity levels are generally lower in the mornings and higher in the afternoons, reflecting daily fluctuations in the temperature. This could be caused by rises in the sun's temperature and cooling of the Earth's surface.

#### Exporting Scatterplot

Let's now save the scatterplot depicting temperature patterns by time and hour to my project directory.

```{r exporting scatterplot}
ggsave("temp_thour_sctplot.pdf", height = 4, width = 6)
```

# Problem 2

#### Dataframe for Random Sample & Vectors

Let's create a dataframe that consists of a random sample size of with a standard normal distribution. This will also include the following:

* A logical vector indicating whether elements in that sample are greater than 0
* A character vector of length 10
* A factor vector length 10 depicting 3 different factor levels

```{r dataframe for random sample}
set.seed(123)

rsamp_df = tibble(
  norm_rsamp = rnorm(10, mean = 0, sd = 1),
  gt_zero = norm_rsamp > 0,
  char_vec = as.character(letters[1:10]),
  factor_vec = factor(
    sample(c("Level 1", "Level 2", "Level 3"), 10, TRUE),
  levels = c("Level 1", "Level 2", "Level 3")
  )
)
```

We will now take the mean for each variable listed in the random sample dataframe (i.e., `rsamp_df`) listed above.`

```{r collapse = TRUE}
rsamp_df |> 
  pull(norm_rsamp) |> 
  mean()

rsamp_df |> 
  pull(gt_zero) |> 
  mean()

rsamp_df |> 
  pull(char_vec) |> 
  mean()

rsamp_df |> 
  pull(factor_vec) |> 
  mean()
```

Based on these results, we are able to get the mean for the variables `norm_rsamp` and `gt_zero`. This is beause `norm_rsamp` is averaging each random, numeric observation from the sample, whereas R is coercing **TRUE** and **FALSE** values in `gt_zero` as 1 and 0, respectively, which allows us to make operations such as the mean.

However, we are unable to calculate the mean for the variables `char_vec` and `factor_vec`.  This is because `char_vec` is treating the data as text. Even if its observations were numbers, R will treat all contents in the character vector as texts. In the case `factor_vec`, it is encoding a vector as categorical data, thus lacking numerical value to get the mean or statistical validity from numbers that are assigned to categories.


#### Coercion of Different Data Types

We will now attempt to convert all vector variables in my dataframe, `rsamp_df`, to numeric values using the `as.numerci` function.

```{r eval = FALSE}
rsamp_df |> 
  mutate(
    gt_zero_an = gt_zero |> 
      as.numeric(),
    char_vec_an = char_vec |> 
      as.numeric(),
    factor_vec_an = factor_vec |> 
      as.numeric()
  )
```
