---
title: "Homework 1"
author: "Sergio Ozoria Ramírez (smo2147)"
date: "2025-09-18"
output: github_document
---

# Initial Setup

Loading the tidyverse package for scatterplot + moderndive package to access the penguins dataset.

```{r initial setup, message = FALSE}
library(tidyverse)
library(moderndive)
```

# Problem 1

#### Description of Dataset

We will now access the penguins dataset and provide a short description of the data using inline R code.

```{r describing dataset}
data("early_january_weather")
```

The penguins dataset has **`r nrow(early_january_weather)` rows** and **`r ncol(early_january_weather)` columns**. One important variable that stands out in this dataset is the **temperature** variable (i.e., `temp`), along with hour (i.e., `hour`) and the wind speed (i.e., `wind_speed`), which are critical in regulating/influencing temperature levels. It is important to note that the average temperature has a mean of **`r early_january_weather |> pull(temp) |> mean()`** in Fahrenheit. Moreover, the average hour is **`r early_january_weather |> pull(hour) |> mean()`** and the average wind speed is **`r early_january_weather |> pull(wind_speed) |> mean()`**.

Here's a list of the remaining variables in the dataset:

* `origin`
* `year`
* `month`
* `hour`
* `day`
* `dewp`
* `humid`
* `wind_dir`
* `wind_gust`
* `precip`
* `pressure`
* `visib`
* `time_hour`

#### Dataframe for Scatterplot

```{r}
weather_df =
  early_january_weather |>
  select(time_hour, temp, humid
  )
```

#### Plotting Scatterplot

We will now plot the variables `temp` and `time_hour` to see how the temperature changed by the hour.

```{r}
ggplot(
  weather_df, aes
       (x = time_hour, 
        y = temp, 
        color = humid)) + 
  geom_point() + 
  labs(
    x = "Hour of the Day", 
    y = "Temperature (°F)", 
    color = "Humidity (%)") +
  ggtitle("Temperature by Hour") +
  theme(plot.title = element_text(hjust = 0.25))
```

